p-adic precision

### 2014-01-29 padicprec.tex
Small fixes and changes

The second approach consists in remarking that $f \circ f$ maps
$(x,y)$ to $(2x, 2y)$. CComputing $f \circ f(x,y)$ using the latter
formula, it is clear that the result we get is known up to precision
$(O(p^n), O(p^m))$ --- and even $(O(p^{n+1}), O(p^{m+1}))$ if $p$
equals $2$.

@@

Let us first recall that a square matrix $M$ with coefficients in $\Qp$ 
admits a LU factorization if it can be written as a produit $LU$ where 
$L$ and $U$ is lower triangular and upper triangular respectively. 
The computation of a LU factorization appears as an important tool to 
tackle many classical questions about matrices or linear systems.

@@

An \emph{absolute value} on a field $K$ is a multiplicative map $|\cdot| 
: K \to \R_+$ satisfying the triangle inequality and sending every 
nonzero element of $K$ to a strictly positive real number. It is said 
\emph{ultrametric} if it satisfies the following strong from of the 
triangle inequality:
$$|x+y| \leq \max(|x|, |y|).$$
Given an absolute value $|\cdot|$ on $K$, we note that $|K^\star|$ is a 
multiplicative subgroup of $\R_+^\star$. It is then either discrete or 
dense. We shall say that $|\cdot|$ is \emph{discrete} if $|K^\star|$ is
discrete. The valued field $(K, |\cdot|)$ is said \emph{complete} if all 
Cauchy sequences (with respect to the distance $d$ defined by $d(x,y) = 
|x-y|$) converge.

@@

\begin{deftn}
Let $V, W$ be two normed $K$-vector spaces, let $U \subset V$ be an open 
subset of $V$ and let $f : U \rightarrow W$ be a map. Then $f$ is called 
\emph{differentiable} at $v_0 \in U$ if there exists a 
continuous linear map $f'(v_0) : U \rightarrow W$ such that for any 
$\varepsilon >0$, there exists an open neighbourhoud $U_\varepsilon 
\subset U$ containing $v_0$ with:
\[ 
\forall v, w \in U_\varepsilon, \quad
\Vert f(v)-f(w)-f'(v_0) \cdot \left( v-w \right) \Vert 
\leq \varepsilon \Vert v-w \Vert. 
\]
The linear map $f'(v_0)$ is called the \emph{differential} of $f$ at $v_0$.
\end{deftn}

@@

The first part of Lemma \ref{lem:reductionball} asserts that the 
identity map $\iota_H : (E, \Vert \cdot \Vert_H) \to (E, \Vert \cdot 
\Vert)$ is an homeomorphism. Furthermore, by the second of the Lemma, 
$H$ is the image under $\iota_H$ of a lattice satisfying an extra 
assuption: it is ``between the open unit ball and the closed unit 
ball''.

@@

Let $r \in (0, \delta)$.
Let us show that $f$ maps surjectively $H$ onto $f'(0) \cdot H$.
Let $x \in H$. We must show that $f(x) \in f'(0) \cdot H$. 
By differentiability at $0$, $\Vert f(x)-f'(0) \cdot x \Vert \leq 
\varepsilon \Vert x \Vert $. Therefore, there exists $y \in f'(0) \cdot 
H$ such that 
$\Vert y \Vert \leq \varepsilon r$. By the open mapping Theorem and the 
surjectify of $f'(0)$, there exists $x_1 \in E$ such that $f'(0) \cdot 
x_1 =y$ and $\Vert x_1 \Vert \leq \varepsilon C \cdot r < r$. Thus, $x_1 
\in B^-_E(r) \subset H$ and $f(x)= f'(0) \cdot (x-x_1) \in f'(0) 
\cdot H$.

We shall now prove surjectivity. Let $y \in f'(0) \cdot H$. 
Let $x_0 \in H$ be such that $y = f'(0) \cdot x_0$.
We define by induction two sequences $(x_n)$ and $(z_n)$ by
\begin{itemize}
\item $z_n \in E$ is such that
$f'(0) \cdot z_n = y - f(x_n)$ and
$\Vert z_n \Vert \leq C \cdot \Vert y - f(x_n) \Vert$,
\item $x_{n+1}=x_n+z_n$.
\end{itemize}
By definition, $\Vert z_0 \Vert < C \Vert y_0\Vert$ and $y_0 
= -f(x_0)+f'(0) \cdot x_0$. Therefore, $\Vert y_0 \Vert < \varepsilon 
\Vert x_0 \Vert \leq \varepsilon r$. Thus $\Vert z_0 \Vert \leq 
\varepsilon C \cdot r < r$. Thus, $x_1$ and $z_0$ lie in $B^-_E(r) 
\subset H$. An easy induction involving an analoguous computation
shows that the sequences $(x_n)$ and $(z_n)$ are well defined and
take their values in $H$.

@@

Before discussing applications of Lemma \ref{lem:main} to ultrametric 
precision, we would like to describe a little bit more restrictive 
setting where the constant $\delta$ appearing in this Lemma can be 
computed explicitely.
We refer to Schneider \cite{Schneider}, \S 6, for a definition of a 
locally analytic function.

\begin{rmk}
Let $f$ be locally analytic function $K^n \rightarrow K^m$. Let $x \in K^n$ be such that $f'(x)$ is surjective. Then the radius of the ball needed in can be read on the Newton polygon given by the norms of the Taylor serie expansion of $f$ at $x$. We assume the $C$ given by the Banach-Schauder is given, $\varepsilon = \frac{1}{C}$ and now, we try to recover the link between $\varepsilon$ and $\delta$.
\end{rmk}

\begin{proof}
Let us assume that $x=0$, $f(0)=0$, and for $r>0$, $f(x)=f'(0) \cdot x +\sum_{n \geq 2} \frac{f^{(n)}(0)}{n!} \cdot (x, \dots,x)$ is the Taylor serie expansion of $f$ at $0$, coinciding on $B_E (0,r)$.

@@

By Legendre formula, it is enough that for all $n \geq 2$:
\[\log_p (\interleave f^{(n)}(0) \interleave )-(n-1) (\log_p l-(n-1)) \geq - \log_p \varepsilon.\]

If we denote by $NP_F$ the Newton polygon application of the power serie $F(t) = \sum_{k\geq 0}^{+\infty} \log_p (\interleave f^{(k+1)}(0) \interleave ) t^k$, then it is enough that:
\[ \sup_{ x \in \mathbb{R}_+} \left( NP_F(x) -x  \left( \log_p l - \frac{1}{p-1} \right) \right) \geq - \log_p \varepsilon.\]

@@

It may happen that natural $p$-adic ``objects'' do not lie in vector 
spaces. Actually, examples are numerous: points on the projective space 
or elliptic curves, subspaces of a fixed vector space (which lie in 
grassmannians), classes of isomorphism of certain curves (which lie in 
various moduli spaces), \emph{etc.} 
That is why we think that it is quite important to extend the formalism 
developped in Section \ref{sec:philosophy} to a more general setting. 
In this Section, we consider the quite general case of a strictly
differentiable manifolds locally modeled on (eventually infinite 
dimensional) ultrametric Banach spaces. This covers all the 
aforementioned examples.

@@

\begin{deftn}
Let $x \in V$. 
A \emph{precision data} at $x$ is a lattice in the tangent space 
$T_x V$.
\end{deftn}

\begin{deftn}
Let $x \in V$ and $H$ be a precision data at $x$.
For all $i \in I$ such that $x \in V_i$, we set:
$$x + O_i(H) = \varphi_i^{-1}\big(\varphi_i(x) + \varphi_i'(x)(H)\big) 
\subset V.$$
\end{deftn}

@@

\begin{prop}
\label{prop:independance}
Let $x \in V$ and $H$ be a precision data at $x$.
If $H$ is ``small enough'' \todo{(precise!)}, the subset $x + 
O_i(H)$ does not depend on $i$.
\end{prop}

@@

Under the assumption of Proposition \ref{prop:independance}, we define
$x + O(H)$ as $x + O_i(H)$ for some (equivalently all) $i$ such that
$x \in V_i$.

@@

In order to restrict ourselves to elements $x$ lying in a dense 
computable subset, we need to compare $x + O(H)$ with some $x' + 
O(H')$ when $x$ and $x'$ are close enough.
Let us first examine the situation in a fixed given chart: we fix some 
index $i \in I$ and pick two elements $x$ and $x'$ in $V_i$. We consider 
in addition a precision data $H$ at $x$ and we want to produce a 
precision data $H'$ at $x'$ such that $x + O_i(H) = x' + O_i(H')$. 
We remark that the tangent spaces $T_x V$ and $T_{x'} V$ are both 
isomorphic to $E_i$ \emph{via} the maps $\varphi'_i(x)$ and 
$\varphi'_i(x')$ respectively. A natural candidate for $H'$ is then:
\begin{equation}
\label{eq:Hprime}
H' = \big(\varphi'_i(x')^{-1} \circ \varphi'_i(x)\big) (H).
\end{equation}
And indeed, one can check that $x + O(H)$ is indeed equal to $x' + 
O(H')$ as soon as $x$ and $x'$ are close enough in the following sense: 
the difference $\varphi_i(x) - \varphi_i(x')$ lies in the lattice $H_i = 
\varphi'_i(x)(H)$. We furthermore have a proprety of independance on 
$i$.

\begin{prop}
We assume that $H$ is ``small enough'' \todo{(precise!)} and that $x$ and 
$x'$ are close enough. Then:
\begin{enumerate}[(i)]
\item the lattice $H'$ defined by Eq.~\eqref{eq:Hprime} does not depend 
on $i$;
\item the couples $(x,H)$ and $(x', H')$ satisfy the assumptions of
Proposition \ref{prop:independance} and we have $x + O(H) = x' + O(H')$.
\end{enumerate}
\end{prop}

\begin{proof}
We first prove (i). For an index $i$ such that $x, x' \in V_i$, let us 
denote by $f_i : T_x V \to T_{x'} V$ the composite $\varphi'_i(x')^{-1} 
\circ \varphi'_i(x)$. Given an extra index $j$ satisfying the same
assumption, we know by \todo{(find a reference)} that the difference
$f_i - f_j$ goes to $0$ when $x'$ converges to $x$. Since $H$ is open
in $T_x V$, this implies that $(f_j - f_i)(H)$ contains $f_i(H)$ and 
$f_j(H)$ if $x'$ and $x$ are close enough. Now, pick $w \in f_j(H)$ and 
write it $w = f_j(v)$ with $v \in H$. Then $w$ is equal to $f_i(v) + 
(f_j - f_i)(v)$ and then belongs to $f_i(H)$ because each summand does. 
Therefore $f_j(H) \subset f_i(H)$. The inverse inclusion can be proved 
in the same way.

@@

In this last paragraph, we illustrate the theory developed in this 
article by designing an algorithm for computing the $u_n$'s at precision 
$O(p^N)$. The algorithm is presented at the top of page 
\pageref{algo:somos}. We now prove that it is correct.
%
\begin{algorithm}[t]
\SetKwInOut{Assumption}{Assumption}
\KwIn{$a,b,c,d$ --- four initial terms of a SOMOS 4 sequence $(u_n)_{n \geq 0}$}
\KwIn{$n, N$ --- two integers}
\Assumption{$a$, $b$, $c$ and $d$ lie in $\Z_p^\times$ and are known at 
precision $O(p^N)$}
\Assumption{None of $u_i$ $(0 \leq i \leq n$) is divisible by $p^N$}
\KwOut{$u_n$ at precision $O(p^N)$}
\BlankLine
$\prec$ $\leftarrow$ $N$\;
\For {$i$ from $1$ to $n-3$}{
  $\prec$ $\leftarrow$ $\prec + v_p(bd + c^2) - v_p(a)$\;
  {\bf lift} (arbitrary) 
    $b$, $c$ and $d$ to precision $O(p^{\prec})$\;
  $\prec$ $\leftarrow$ $\prec - v_p(a)$\;
  $e$ $\leftarrow$ $\frac{bd+c^2} a$; \hspace{1cm}
   \tcp{$e$ is known at precision $O(p^\prec)$}
  $a, b, c, d$ $\leftarrow$ $b + O(p^{\prec}), c + O(p^{\prec}), d + O(p^{\prec}), e$\;
}
\Return{$d + O(p^N)$}\;
\caption{\sc SOMOS$(a, b, c, d, n, N)$}\label{algo:somos}
\end{algorithm}
%
We introduce the (partial) function $f : \Q_p^4 \to \Q_p^4$ defined by 
$f(a,b,c,d) = (b,c,d,\frac{bd+c^2}a)$, so that, for all $i$, we have 
$(u_i, u_{i+1}, u_{i+2}, u_{i+3}) = f_i(u_0, u_1, u_2, u_3)$ where $f_i
= f \circ \cdots \circ f$ ($i$ times). Apparently $f$ is strictly 
differentiable on $\Q_p^\star \times \Q_p^3$ and its differential in the 
canonical basis is given by the matrix:
$$D(a,b,c,d) = \begin{pmatrix}
0 & 1 & 0 & 0 \\
0 & 0 & 1 & 0 \\
0 & 0 & 0 & 1 \\
-\frac{bd+c^2}{a^2} & \frac d a & \frac {2c} a & \frac b a
\end{pmatrix}$$
whose determinant is $\frac{bd+c^2}{a^2}$. Thus, if the $(i+4)$-th term 
of the SOMOS sequence is defined, the mapping $f_i$ is differentiable
at $(u_0, u_1, u_2, u_3)$ and the matrix its differential at this point 
is:
$$D_i = D(u_{i-1}, u_i, u_{i+1}, u_{i+2}) \cdots
D(u_1, u_2, u_3, u_4) \cdot D(u_0, u_1, u_2, u_3).$$
%Its determinant is:
%$$\det D_i = \frac{u_{i+3}}{u_{i-1}} \cdots 
%\frac{u_5}{u_1} \cdot \frac{u_4}{u_0} =
%\frac{u_i \cdot u_{i+1} \cdot u_{i+2} \cdot u_{i+3}}
%{u_0 \cdot u_1 \cdot u_2 \cdot u_3}.$$
Moreover thanks to the Laurent phenomenon, we know that $D_i$ has
integral coefficients. Hence, letting $\varphi_i = f'_i(u_0, u_1, u_2, 
u_3)$, we get 
$$(\det D_i) \cdot p^N \Z_p^4 \subset \varphi_i(p^N \Z_p^4) 
\subset p^N \Z_p^4$$
for all integer $N$.

>>>>

The second approach consists in remarking that $f \circ f$ maps
$(x,y)$ to $(2x, 2y)$.  Computing $f \circ f(x,y)$ using the latter
formula, it is clear that the result we get is known up to precision
$(O(p^n), O(p^m))$ --- and even $(O(p^{n+1}), O(p^{m+1}))$ if $p$
equals $2$.

@@

Let us first recall that a square matrix $M$ with coefficients in $\Qp$ 
admits a LU factorization if it can be written as a product $LU$ where 
$L$ and $U$ is lower triangular and upper triangular respectively. 
The computation of a LU factorization appears as an important tool to 
tackle many classical questions about matrices or linear systems.

@@

An \emph{absolute value} on a field $K$ is a multiplicative map $|\cdot| 
: K \to \R_+$ satisfying the triangle inequality and sending every 
nonzero element of $K$ to a strictly positive real number. It is said 
\emph{ultrametric} if it satisfies the following strong from of the 
triangle inequality:
$$|x+y| \leq \max(|x|, |y|).$$
Given an absolute value $|\cdot|$ on $K$, we note that $|K^\times|$ is a 
multiplicative subgroup of $\R_+^\times$. It is then either discrete or 
dense. We shall say that $|\cdot|$ is \emph{discrete} if $|K^\times|$ is
discrete. The valued field $(K, |\cdot|)$ is said \emph{complete} if all 
Cauchy sequences (with respect to the distance $d$ defined by $d(x,y) = 
|x-y|$) converge.

@@

\begin{deftn}
Let $V, W$ be two normed $K$-vector spaces, let $U \subset V$ be an open 
subset of $V$ and let $f : U \rightarrow W$ be a map. Then $f$ is called 
\emph{differentiable} at $v_0 \in U$ if there exists a 
continuous linear map $f'(v_0) : U \rightarrow W$ such that for any 
$\varepsilon >0$, there exists an open neighborhood $U_\varepsilon 
\subset U$ containing $v_0$ with:
\[ 
\forall v, w \in U_\varepsilon, \quad
\Vert f(v)-f(w)-f'(v_0) \cdot \left( v-w \right) \Vert 
\leq \varepsilon \Vert v-w \Vert. 
\]
The linear map $f'(v_0)$ is called the \emph{differential} of $f$ at $v_0$.
\end{deftn}

@@

The first part of Lemma \ref{lem:reductionball} asserts that the 
identity map $\iota_H : (E, \Vert \cdot \Vert_H) \to (E, \Vert \cdot 
\Vert)$ is an homeomorphism. Furthermore, by the second of the Lemma, 
$H$ is the image under $\iota_H$ of a lattice satisfying an extra 
assumption: it is ``between the open unit ball and the closed unit 
ball''.

@@

Let $r \in (0, \delta)$.
Let us show that $f$ maps surjectively $H$ onto $f'(0) \cdot H$.
Let $x \in H$. We must show that $f(x) \in f'(0) \cdot H$. 
By differentiability at $0$, $\Vert f(x)-f'(0) \cdot x \Vert \leq 
\varepsilon \Vert x \Vert $. Therefore, there exists $y \in f'(0) \cdot 
H$ such that 
$\Vert y \Vert \leq \varepsilon r$. By the open mapping Theorem and the 
surjectivityy of $f'(0)$, there exists $x_1 \in E$ such that $f'(0) \cdot 
x_1 =y$ and $\Vert x_1 \Vert \leq \varepsilon C \cdot r < r$. Thus, $x_1 
\in B^-_E(r) \subset H$ and $f(x)= f'(0) \cdot (x-x_1) \in f'(0) 
\cdot H$.

We shall now prove surjectivity. Let $y \in f'(0) \cdot H$. 
Let $x_0 \in H$ be such that $y = f'(0) \cdot x_0$.
We define by induction two sequences $(x_n)$ and $(z_n)$ by
\begin{itemize}
\item $z_n \in E$ is such that
$f'(0) \cdot z_n = y - f(x_n)$ and
$\Vert z_n \Vert \leq C \cdot \Vert y - f(x_n) \Vert$,
\item $x_{n+1}=x_n+z_n$.
\end{itemize}
By definition, $\Vert z_0 \Vert < C \Vert y_0\Vert$ and $y_0 
= -f(x_0)+f'(0) \cdot x_0$. Therefore, $\Vert y_0 \Vert < \varepsilon 
\Vert x_0 \Vert \leq \varepsilon r$. Thus $\Vert z_0 \Vert \leq 
\varepsilon C \cdot r < r$. Thus, $x_1$ and $z_0$ lie in $B^-_E(r) 
\subset H$. An easy induction involving an analogous computation
shows that the sequences $(x_n)$ and $(z_n)$ are well defined and
take their values in $H$.

@@

Before discussing applications of Lemma \ref{lem:main} to ultrametric 
precision, we would like to describe a little bit more restrictive 
setting where the constant $\delta$ appearing in this Lemma can be 
computed explicitly.
We refer to Schneider \cite{Schneider}, \S 6, for a definition of a 
locally analytic function.

\begin{rmk}
Let $f$ be locally analytic function $K^n \rightarrow K^m$. Let $x \in K^n$ be such that $f'(x)$ is surjective. Then the radius of the ball needed can be read from the Newton polygon given by the norms of the Taylor series expansion of $f$ at $x$. We assume the $C$ given by the Banach-Schauder is given, $\varepsilon = \frac{1}{C}$ and now, we try to recover the link between $\varepsilon$ and $\delta$.
\end{rmk}

\begin{proof}
Let us assume that $x=0$, $f(0)=0$, and for $r>0$, $f(x)=f'(0) \cdot x +\sum_{n \geq 2} \frac{f^{(n)}(0)}{n!} \cdot (x, \dots,x)$ is the Taylor series expansion of $f$ at $0$, coinciding on $B_E (0,r)$.

@@

By Legendre formula, it is enough that for all $n \geq 2$:
\[
\log_p (\interleave f^{(n)}(0) \interleave )-(n-1) (\log_p l-(n-1)) \geq - \log_p \varepsilon.
\]

If we denote by $NP_F$ the Newton polygon application of the power series 
\[
F(t) = \sum_{k\geq 0}^{+\infty} \log_p (\interleave f^{(k+1)}(0) \interleave ) t^k
\]
then it is enough that:
\[
\sup_{ x \in \mathbb{R}_+} \left( NP_F(x) -x  \left( \log_p l - \frac{1}{p-1} \right) \right) \geq - \log_p \varepsilon.
\]

@@

Many natural $p$-adic objects do not lie in vector spaces:
points in projective spaces or elliptic curves,
subspaces of a fixed vector space (which lie in Grassmannians),
classes of isomorphism of certain curves (which lie in various moduli spaces),
\emph{etc.}  In this section we extend the formalism 
developed in Section \ref{sec:philosophy} to a more general setting:
we consider the quite general case of strictly
differentiable manifolds locally modeled on ultrametric Banach spaces.
This covers all the 
aforementioned examples.

@@

\begin{deftn}
Let $x \in V$. 
A \emph{precision datum} at $x$ is a lattice in the tangent space 
$T_x V$.
\end{deftn}

\begin{deftn}
Let $x \in V$ and $H$ be a precision datum at $x$.
For all $i \in I$ such that $x \in V_i$, we set:
$$x + O_i(H) = \varphi_i^{-1}\big(\varphi_i(x) + \varphi_i'(x)(H)\big) 
\subset V.$$
\end{deftn}

@@

\begin{prop}
\label{prop:independence}
Let $x \in V$ and $H$ be a precision data at $x$.
If $H$ is ``small enough'' \todo{(precise!)}, the subset $x + 
O_i(H)$ does not depend on $i$.
\end{prop}

@@

Under the assumption of Proposition \ref{prop:independence}, we define
$x + O(H)$ as $x + O_i(H)$ for some (equivalently all) $i$ such that
$x \in V_i$.

@@

In order to restrict ourselves to elements $x$ lying in a dense 
computable subset, we need to compare $x_0 + O(H_0)$ with varying $x + 
O(H)$ when $x$ and $x_0$ are close enough.
Let us first examine the situation in a fixed given chart: we fix some 
index $i \in I$ and pick two elements $x_0$ and $x$ in $V_i$. We consider 
in addition a precision datum $H_0$ at $x_0$ and we want to produce a 
precision datum $H$ at $x$ such that $x_0 + O_i(H_0) = x + O_i(H)$. 
We remark that the tangent spaces $T_{x_0} V$ and $T_x V$ are both 
isomorphic to $E_i$ via the maps $\varphi'_i(x_0)$ and 
$\varphi'_i(x)$ respectively. A natural candidate for $H$ is then:
\begin{equation}
\label{eq:Hprime}
H = \left(\varphi'_i(x)^{-1} \circ \varphi'_i(x_0)\right) (H_0).
\end{equation}
And indeed, one can check that $x + O(H)$ is indeed equal to $x_0 + 
O(H_0)$ as soon as $x$ and $x_0$ are close enough in the following sense: 
the difference $\varphi_i(x) - \varphi_i(x_0)$ lies in the lattice $H_i = 
\varphi'_i(x_0)(H_0)$. We furthermore have a property of independence on 
$i$.

\begin{prop}
We assume that $H_0$ is ``small enough'' \todo{(precise!)} and that $x$ is close 
enough to $x_0$. Then
\begin{enumerate}[(i)]
\item the lattice $H$ defined by \eqref{eq:Hprime} does not depend 
on $i$,
\item the pairs $(x,H)$ and $(x_0, H_0)$ satisfy the assumptions of
Proposition \ref{prop:independence} and we have $x + O(H) = x_0 + O(H_0)$.
\end{enumerate}
\end{prop}

\begin{proof}
We first prove (i). For an index $i$ such that $x, x_0 \in V_i$, let us 
denote by $f_i : T_{x_0} V \to T_x V$ the composite $\varphi'_i(x)^{-1} 
\circ \varphi'_i(x_0)$. Given an extra index $j$ satisfying the same
assumption, we know by \todo{(find a reference)} that the difference
$f_i - f_j$ goes to $0$ when $x$ converges to $x_0$. Since $H_0$ is open
in $T_{x_0} V$, this implies that $(f_j - f_i)(H_0)$ contains $f_i(H_0)$ and 
$f_j(H_0)$ if $x$ and $x_0$ are close enough. Now, pick $w \in f_j(H_0)$ and 
write it $w = f_j(v)$ with $v \in H_0$. Then $w$ is equal to $f_i(v) + 
(f_j - f_i)(v)$ and then belongs to $f_i(H_0)$ because each summand does. 
Therefore $f_j(H_0) \subset f_i(H_0)$. The inverse inclusion can be proved 
in the same way.

@@

In this last paragraph, we illustrate the theory developed in this 
article by designing an algorithm for computing the $u_n$'s at precision 
$O(p^N)$. The algorithm is presented at the top of page 
\pageref{algo:somos}. We now prove that it is correct.
%
\begin{algorithm}[t]
\SetKwInOut{Assumption}{Assumption}
\KwIn{$a,b,c,d$ --- four initial terms of a SOMOS 4 sequence $(u_n)_{n \geq 0}$}
\KwIn{$n, N$ --- two integers}
\Assumption{$a$, $b$, $c$ and $d$ lie in $\Z_p^\times$ and are known at 
precision $O(p^N)$}
\Assumption{None of $u_i$ $(0 \leq i \leq n$) is divisible by $p^N$}
\KwOut{$u_n$ at precision $O(p^N)$}
\BlankLine
$\prec$ $\leftarrow$ $N$\;
\For {$i$ from $1$ to $n-3$}{
  $\prec$ $\leftarrow$ $\prec + v_p(bd + c^2) - v_p(a)$\;
  {\bf lift} (arbitrary) 
    $b$, $c$ and $d$ to precision $O(p^{\prec})$\;
  $\prec$ $\leftarrow$ $\prec - v_p(a)$\;
  $e$ $\leftarrow$ $\frac{bd+c^2} a$; \hspace{1cm}
   \tcp{$e$ is known at precision $O(p^\prec)$}
  $a, b, c, d$ $\leftarrow$ $b + O(p^{\prec}), c + O(p^{\prec}), d + O(p^{\prec}), e$\;
}
\Return{$d + O(p^N)$}\;
\caption{\sc SOMOS$(a, b, c, d, n, N)$}\label{algo:somos}
\end{algorithm}
%
We introduce the (partial) function $f : \Q_p^4 \to \Q_p^4$ defined by 
$f(a,b,c,d) = (b,c,d,\frac{bd+c^2}a)$, so that, for all $i$, we have 
$(u_i, u_{i+1}, u_{i+2}, u_{i+3}) = f_i(u_0, u_1, u_2, u_3)$ where $f_i
= f \circ \cdots \circ f$ ($i$ times). Clearly, $f$ is strictly 
differentiable on $\Q_p^\times \times \Q_p^3$ and its differential in the 
canonical basis is given by the matrix:
$$D(a,b,c,d) = \begin{pmatrix}
0 & 1 & 0 & 0 \\
0 & 0 & 1 & 0 \\
0 & 0 & 0 & 1 \\
-\frac{bd+c^2}{a^2} & \frac d a & \frac {2c} a & \frac b a
\end{pmatrix}$$
whose determinant is $\frac{bd+c^2}{a^2}$. Thus, if the $(i+4)$-th term 
of the SOMOS sequence is defined, the mapping $f_i$ is differentiable
at $(u_0, u_1, u_2, u_3)$ and the matrix its differential at this point 
is:
$$D_i = D(u_{i-1}, u_i, u_{i+1}, u_{i+2}) \cdots
D(u_1, u_2, u_3, u_4) \cdot D(u_0, u_1, u_2, u_3).$$
%Its determinant is:
%$$\det D_i = \frac{u_{i+3}}{u_{i-1}} \cdots 
%\frac{u_5}{u_1} \cdot \frac{u_4}{u_0} =
%\frac{u_i \cdot u_{i+1} \cdot u_{i+2} \cdot u_{i+3}}
%{u_0 \cdot u_1 \cdot u_2 \cdot u_3}.$$
Moreover thanks to the Laurent phenomenon, we know that $D_i$ has
integral coefficients. Hence, letting $\varphi_i = f'_i(u_0, u_1, u_2, 
u_3)$, we get 
$$(\det D_i) \cdot p^N \Z_p^4 \subset \varphi_i(p^N \Z_p^4) 
\subset p^N \Z_p^4$$
for all integer $N$.
